### 2021-05-24 concurrency/docs/theory/theory.tex
continue working out rwlock hoare proofs

Second, we have to show why it is allowed to use borrowed state to justify the transitions.
Luckily, from (\ref{invTotal}) we can derive this law,
\[
  \kappa \ltGe active \implies View_\kappa(\gamma)\left(
      \left(
        \Bigcircle_{\kappa' \ltGe active} reserved(\kappa,\gamma)
      \right) \cdot
      \left(
        \Bigcircle_{\delta} Project_\delta\left(Total_{active}(\delta \refines_{\rel} \gamma)\right)
      \right)
    \right).
    \label{invDerived}
\]
This large expression inside the $View_\kappa(\gamma)$ is just the part of
$Total_{active}(\gamma)$ except the part \emph{which is live at $\gamma$}.
Thus, combined with \ref{invView}, we can see that any borrow will be covered by the reserved 
state at $\gamma$ plus all state in the refined locations.

@@

These laws are for exclusive locks:
\begin{align*}
\dboxed{\Central\left(\false, rc, x\right)}^{\alpha \refines \gamma} &\viewshift \dboxed{\Central\left(\true, rc, x\right)}^{\alpha \refines \gamma} * \dboxed{\ExcPending}^{\alpha \refines \gamma}\\
%\Central\left(exc, 0, x\right) \cdot \ExcPending &\mapsto \Central\left(exc, 0, \bot\right) \cdot \ExcGuard
\dboxed{\Central\left(exc, 0, x\right)}^{\alpha \refines \gamma} * \dboxed{\ExcPending}^{\alpha \refines \gamma}
    &\viewshift
    \dboxed{\Central\left(exc, 0, x\right)}^{\alpha \refines \gamma} * \dboxed{\ExcGuard}^{\alpha \refines \gamma} * \dboxed{x}^\gamma\\
\dboxed{\Central\left(exc, rc, y\right)}^{\alpha \refines \gamma} * \dboxed{\ExcGuard}^{\alpha \refines \gamma} * \dboxed{x}^\gamma
    &\viewshift
        \dboxed{\Central\left(\false, rc, x\right)}^{\alpha \refines \gamma}\\
\end{align*}
The first two are for acquisition, while the third is for release. Meanwhile,
the first one is derived from Ghost-ViewShift, while the last two are derived from
Ghost-Refinement-Exchange.

These laws are for shared locks:
\begin{align*}
\dboxed{\Central\left(exc, rc, x\right)}^{\alpha \refines \gamma} &\viewshift
  \dboxed{\Central\left(exc, rc+1, x\right)}^{\alpha \refines \gamma} * \dboxed{\SharedPending}^{\alpha \refines \gamma}\\
\dboxed{\Central\left(false, rc, x\right)}^{\alpha \refines \gamma} * \dboxed{\SharedPending}^{\alpha \refines \gamma}
    &\viewshift
      \dboxed{\Central\left(false, rc, x\right)}^{\alpha \refines \gamma} * \dboxed{\SharedGuard(x)}^{\alpha \refines \gamma}\\
\dboxed{\Central\left(exc, rc, x\right)}^{\alpha \refines \gamma} * \dboxed{\SharedGuard(x)}^{\alpha \refines \gamma}
    &\viewshift
        \dboxed{\Central\left(exc, rc - 1, x\right)}^{\alpha \refines \gamma}
\end{align*}
All of these are derived from Ghost-ViewShift.

@@

>>>>

Second, we have to show why it is allowed to use borrowed state to justify the transitions.
Luckily, from (\ref{invTotal}) we can derive this law,
\[
  \kappa \ltGe active \implies View_\kappa(\gamma)\left(
      \left(
        \Bigcircle_{\kappa' \ltGe active} reserved(\kappa,\gamma)
      \right) \cdot
      \left(
        \Bigcircle_{\delta} Project_\delta\left(Total_{active}(\delta \refines_{\rel} \gamma)\right)
      \right)
    \right).
    \label{invDerived}
\]
This large expression inside the $View_\kappa(\gamma)$ is just the part of
$Total_{active}(\gamma)$ except the part \emph{which is live at $\gamma$}.
Thus, combined with (\ref{invView}), we can see that any borrow will be covered by the reserved 
state at $\gamma$ plus all state in the refined locations.

@@

These laws are for exclusive locks:
\begin{align*}
\dboxed{\Central\left(\false, rc, x\right)}^{\alpha \refines \gamma} &\viewshift \dboxed{\Central\left(\true, rc, x\right)}^{\alpha \refines \gamma} * \dboxed{\ExcPending}^{\alpha \refines \gamma} && \text{(exc-begin)} \\
%\Central\left(exc, 0, x\right) \cdot \ExcPending &\mapsto \Central\left(exc, 0, \bot\right) \cdot \ExcGuard
\dboxed{\Central\left(exc, 0, x\right)}^{\alpha \refines \gamma} * \dboxed{\ExcPending}^{\alpha \refines \gamma}
    &\viewshift
    \dboxed{\Central\left(exc, 0, x\right)}^{\alpha \refines \gamma} * \dboxed{\ExcGuard}^{\alpha \refines \gamma} * \dboxed{x}^\gamma && \text{(exc-finish)} \\
\dboxed{\Central\left(exc, rc, y\right)}^{\alpha \refines \gamma} * \dboxed{\ExcGuard}^{\alpha \refines \gamma} * \dboxed{x}^\gamma
    &\viewshift
        \dboxed{\Central\left(\false, rc, x\right)}^{\alpha \refines \gamma} && \text{(exc-return)} \\
\end{align*}
The first two are for acquisition, while the third is for release. Meanwhile,
the first one is derived from Ghost-ViewShift, while the last two are derived from
Ghost-Refinement-Exchange.

These laws are for shared locks:
\begin{align*}
\dboxed{\Central\left(exc, rc, x\right)}^{\alpha \refines \gamma} &\viewshift
  \dboxed{\Central\left(exc, rc+1, x\right)}^{\alpha \refines \gamma} * \dboxed{\SharedPending}^{\alpha \refines \gamma} && \text{(shared-begin)} \\
\dboxed{\Central\left(exc, rc, x\right)}^{\alpha \refines \gamma}  * \dboxed{\SharedPending}^{\alpha \refines \gamma} &\viewshift
  \dboxed{\Central\left(exc, rc-1, x\right)}^{\alpha \refines \gamma} && \text{(shared-abort)} \\
\dboxed{\Central\left(false, rc, x\right)}^{\alpha \refines \gamma} * \dboxed{\SharedPending}^{\alpha \refines \gamma}
    &\viewshift
      \dboxed{\Central\left(false, rc, x\right)}^{\alpha \refines \gamma} * \dboxed{\SharedGuard(x)}^{\alpha \refines \gamma} && \text{(shared-return)}\\
\dboxed{\Central\left(exc, rc, x\right)}^{\alpha \refines \gamma} * \dboxed{\SharedGuard(x)}^{\alpha \refines \gamma}
    &\viewshift
        \dboxed{\Central\left(exc, rc - 1, x\right)}^{\alpha \refines \gamma} && \text{(shared-abort)}
\end{align*}
All of these are derived from Ghost-ViewShift.

@@

\textbf{RwLock Verified Example}

\newcommand{\rwlock}{rwlock}
\newcommand{\exc}{exc}
\newcommand{\rc}{rc}
\newcommand{\Pointer}{Pointer}

\newcommand{\AcquireExc}{\text{AcquireExc}}
\newcommand{\AcquireShared}{\text{AcquireShared}}
%\newcommand{\AcquireExcSetBit}{\text{AcquireExcSetBit}}
%\newcommand{\AcquireExcCheckRc}{\text{AcquireExcCheckRc}}
%\newcommand{\AcquireSharedIncRc}{\text{AcquireSharedIncRc}}
%\newcommand{\AcquireSharedCheckExc}{\text{AcquireSharedCheckExc}}
%\newcommand{\AcquireSharedAbort}{\text{AcquireSharedAbort}}
\newcommand{\ReleaseExc}{\text{ReleaseExc}}
\newcommand{\ReleaseShared}{\text{ReleaseShared}}

\newcommand{\langloop}{\textbf{loop}}
\newcommand{\until}{\textbf{until}}
\newcommand{\langif}{\textbf{if}}
\newcommand{\langthen}{\textbf{then}}
\newcommand{\langelse}{\textbf{else}}

\newcommand{\CAS}{\textbf{CAS}}
\newcommand{\AtomicAdd}{\textbf{AtomicAdd}}

\newcommand{\Inv}{Inv}
\newcommand{\inv}{inv}
\newcommand{\inlinehoare}[1]{{\color{blue}\left\{#1\right\}}}

Let's use a struct $\rwlock := \{\exc: \Pointer, \rc: \Pointer\}$. Our implementation will be straightforward:

\begin{align*}
  \AcquireExc(\rwlock) ~\eqdef~
      & \langloop~\until~\CAS(\rwlock.\exc,0,1);\\
      & \langloop~\until~(!\rwlock.\rc) = 0\\
      \\
  \AcquireShared(\rwlock) ~\eqdef~
      & \langloop~\until~(\\
        & ~~~~ (\langloop~\until~!\rwlock.\exc);\\
        & ~~~~ \AtomicAdd(\rwlock.\rc, 1);\\
        & ~~~~ \langif~ (!\rwlock.\exc) = 0 ~\langthen\\
        & ~~~~~~~~ \true\\
        & ~~~~ \langelse\\
        & ~~~~~~~~ \AtomicAdd(\rwlock.\rc, -1); \false\\
      )\\
      \\
  %\AcquireExcSetBit(\rwlock) ~\eqdef~
  %    & \CAS(\rwlock.\exc,\false,\true)\\
  %    \\
  %\AcquireExcCheckRc(\rwlock) ~\eqdef~
  %    & (!\rwlock.\rc) = 0\\
  %    \\
  %\AcquireSharedIncRc(\rwlock) ~\eqdef~
  %    & \AtomicAdd(\rwlock.\rc, 1)\\
  %    \\
  %\AcquireSharedAbort(\rwlock) ~\eqdef~
  %    & \AtomicAdd(\rwlock.\rc, -1)\\
  %    \\
  %\AcquireSharedCheckExc(\rwlock) ~\eqdef~
  %    & (!\rwlock.\exc) = \false\\
  %    \\
  \ReleaseExc(\rwlock) ~\eqdef~
      & \rwlock.\exc := 0\\
      \\
  \ReleaseShared(\rwlock) ~\eqdef~
      & \AtomicAdd(\rwlock.\rc, -1)
\end{align*}
We would like to prove the following specifications, for some predicate $\Inv(\rwlock,\alpha,\gamma,\inv)$. (Here, $\inv$ is the user-defined predicate on the ghost state stored by the $\rwlock$.)

$\AcquireExc$:
\begin{align*}
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) }\\
& \AcquireExc(\rwlock) \\
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) * \dboxed{\ExcGuard}^{\alpha \refines \gamma} * \exists x.~ \dboxed{x}^\gamma \land \inv(x) }
\end{align*}
$\ReleaseExc$:
\begin{align*}
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) * \dboxed{\ExcGuard}^{\alpha \refines \gamma} * \dboxed{x}^\gamma \land \inv(x) }\\
& \ReleaseExc(\rwlock) \\
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) }
\end{align*}
$\AcquireShared$:
\begin{align*}
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) }\\
& \AcquireShared(\rwlock) \\
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) * \exists x.~ \dboxed{\SharedGuard(x)}^{\alpha \refines \gamma} \land \inv(x) }
\end{align*}
$\ReleaseShared$:
\begin{align*}
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) * \dboxed{\SharedGuard(x)}^{\alpha \refines \gamma} \land \inv(x) }\\
& \ReleaseShared(\rwlock) \\
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) }
\end{align*}
To prove these specifications, we will first define $\Inv$:
\begin{align*}
  \Inv(\rwlock) ~\eqdef~
      \exists \iota, \boxed{
        \exists exc,rc,x .~
            \dboxed{\Central(exc,rc,x)}^{\alpha \refines \gamma}
            * (\rwlock.\exc \pointsto exc)
            * (\rwlock.\rc \pointsto rc)
      }^{~\iota}
\end{align*}
Now, we can work out how to verify the above programs.
$\AcquireExc$ will follow from the below:
\begin{align}
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) } \nonumber \\
& \CAS(\rwlock.\exc,0,1) \nonumber \\ %\label{hoare-cas-exc-pending} \\
& \inlinehoare{v.~
  \Inv(\rwlock,\alpha,\gamma,\inv) *
      \left(v \implies \dboxed{\ExcPending}^{\alpha\refines\gamma}\right) } \nonumber \\
   \nonumber \\
& \inlinehoare{
    \Inv(\rwlock,\alpha,\gamma,\inv) *
    \dboxed{\ExcPending}^{\alpha\refines\gamma} } \nonumber \\
& (!\rwlock.\rc) \nonumber \\
& \inlinehoare{v .~
    \Inv(\rwlock,\alpha,\gamma,\inv) *
    \left(v\ne 0 \implies \dboxed{\ExcPending}^{\alpha\refines\gamma}\right) *
    \left(v=0 \implies \dboxed{\ExcGuard}^{\alpha\refines\gamma} *
        \exists x.~ \dboxed{x}^{\gamma} \land \inv(x)\right)
    } \nonumber
\end{align}
$\AcquireShared$ will follow from:
\begin{align}
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) } \nonumber \\
& \AtomicAdd(\rwlock.\rc, 1) \label{hoare-atomic-add-shared-pending} \\
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) * \SharedPending } \nonumber\\
 \nonumber\\
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) * \SharedPending }\nonumber\\
& (!\rwlock.\exc)\nonumber\\
& \inlinehoare{ v.~
    \Inv(\rwlock,\alpha,\gamma,\inv) * \left(v \ne 0 \implies \SharedPending\right)
        * \left(v = 0 \implies \exists x.~
          \dboxed{\SharedGuard(x)}^{\alpha \refines \gamma} \land \inv(x)\right)
   }\nonumber\\
   \nonumber\\
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) * \SharedPending }\nonumber\\
& \AtomicAdd(\rwlock.\rc, -1)\nonumber\\
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) }\nonumber
\end{align}
$\ReleaseExc$ will follow from:
\begin{align*}
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) * \dboxed{\ExcGuard}^{\alpha \refines \gamma} * \dboxed{x}^\gamma \land \inv(x) }\\
& \rwlock.\exc := 0\\
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) }
\end{align*}
And finally, $\ReleaseShared$ will follow from:
\begin{align*}
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) * \dboxed{\SharedGuard(x)}^{\alpha \refines \gamma} \land \inv(x) }\\
& \AtomicAdd(\rwlock.\rc, -1)\\
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) }
\end{align*}
Let's work out (\ref{hoare-atomic-add-shared-pending}) explicitly:
%\begin{align*}
%& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) } \\
%& \inlinehoare{
%        \exists exc,rc,x .~
%            \dboxed{\Central(exc,rc,x)}^{\alpha \refines \gamma}
%            * (\rwlock.\exc \pointsto exc)
%            * (\rwlock.\rc \pointsto rc)
%    }\\
%& \CAS(\rwlock.\exc,0,1) \\
%& \inlinehoare{v.~
%  \Inv(\rwlock,\alpha,\gamma,\inv) *
%      \left(v \implies \dboxed{\ExcPending}^{\alpha\refines\gamma}\right) }
%\end{align*}
\begin{align*}
& \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) } \\
\text{(open Inv)} ~~~~ & \inlinehoare{
        \exists exc,rc,x .~
            \dboxed{\Central(exc,rc,x)}^{\alpha \refines \gamma}
            * (\rwlock.\exc \pointsto exc)
            * (\rwlock.\rc \pointsto rc)
    }\\
& \AtomicAdd(\rwlock.\rc, 1) \\
& \inlinehoare{
        \exists exc,rc,x .~
            \dboxed{\Central(exc,rc,x)}^{\alpha \refines \gamma}
            * (\rwlock.\exc \pointsto exc)
            * (\rwlock.\rc \pointsto rc + 1)
    }\\
\text{(shared-begin)} ~~~~ & \inlinehoare{
        \exists exc,rc,x .~
            \dboxed{\Central(exc,rc + 1,x)}^{\alpha \refines \gamma}
            * \SharedPending
            * (\rwlock.\exc \pointsto exc)
            * (\rwlock.\rc \pointsto rc + 1)
    }\\
(rc'=rc+1) ~~~~ & \inlinehoare{
        \exists exc,rc',x .~
            \dboxed{\Central(exc,rc',x)}^{\alpha \refines \gamma}
            * \SharedPending
            * (\rwlock.\exc \pointsto exc)
            * (\rwlock.\rc \pointsto rc')
    }\\
\text{(close Inv)} ~~~~ & \inlinehoare{ \Inv(\rwlock,\alpha,\gamma,\inv) * \SharedPending } \nonumber\\
\end{align*}
